Data Munging

### 2025-02-18 manuscript/manuscript_file.Rmd
Completed revisions in response to Lester and Jana's tech review.

**Lakes, ponds, and reservoirs are estimated to be globally important sources of nitrous oxide (N~2~O) to the atmosphere but recent evidence of N~2~O uptake across a broad range of lakes have called the accuracy of emission estimates into question. Here we use a new national-scale dataset on dissolved N~2~O and a Bayesian hierarchical model to predict N~2~O concentration and emission rates in `r sframe %>% summarize(nlake = n()) %>% pull() %>% format(., big.mark=",", scientific=FALSE)` waterbodies in the conterminous U.S. (CONUS). We found that N~2~O undersaturation was widespread through the CONUS, with an estimated `r national_stats$post_m_prop_sat * 100`% (95% credible interval: `r national_stats$LCL_prop_sat * 100` - `r national_stats$UCL_prop_sat * 100`%) of lakes functioning as N~2~O sinks. The model predicts dissolved N~2~O concentrations reasonably well based partly on interactions between nitrate concentration, waterbody size, and water temperature. Despite working with the largest aquatic N~2~O dataset to date, our national-scale estimate of N~2~O emissions from CONUS lakes is poorly constrained, with a 95% credible interval ranging from net uptake to net emission (`r national_stats %>% select(LCL_fn2o) %>% pull() %>% round(., 0)` - `r national_stats %>% select(UCL_fn2o) %>% pull() %>% round(., 0)` metric tons N~2~O year^-1^). Widespread N~2~O undersaturation in CONUS waterbodies and a national-scale emission estimate that is too uncertain to determine if CONUS lakes are a net source or sink of N~2~O highlight the need to revisit N~2~O models which presume surface waters are a N~2~O source.**

@@

During the summer of 2017 the United States Environmental Protection Agency conducted the first national-scale survey of N~2~O concentrations and emissions in CONUS lakes as part of the National Lakes Assessment (NLA) [@USEPA_NLA17_2022].  The population of interest (POI) for the survey was `r sframe %>% summarize(nlake = n()) %>% pull() %>% format(., big.mark=",", scientific=FALSE)` CONUS lakes that had a surface area of at least 1 ha.  A representative sample (n=`r dg %>% filter(sample.source == "DG", visit.no == 1, sitetype == "PROB") %>% nrow()`) was selected from the POI using a generalized random-tessellation stratified survey design [@Stevens_Olsen_2004]**(Fig. 1)**. The survey design was stratified by state and used unequal probability categories based on lake size to ensure that large lakes, which are far less common than small lakes, were included in the study.  Sample sites were distributed across all nine major CONUS ecoregions [@Herlihy_etal_2008] and varied widely in water chemistry, morphometry, watershed land-use, and climate [@epa_data]. The sample data and survey design information (e.g., stratification variables and unequal probability categories) were used to develop a Bayesian hierarchical model to predict dissolved and equilibrium N~2~O concentrations for the POI. Predictor variables for the model were restricted to those required to generate predictions that reproduced the observed N~2~O measures in an out-of-sample dataset and were available for the POI.  See Methods and SI for a detailed discussion of variable selection.

Nearly `r round(((dg %>% dplyr::filter(n2o.src.snk == "sink") %>% {nrow(.)} / dg %>% distinct(site.id) %>% {nrow(.)}) * 100), 0)` percent of the sampled lakes were undersaturated in N~2~O and were therefore functioning as N~2~O sinks **(Fig. 1)**. The Bayesian hierarchical model predicted that the majority lakes in the POI (posterior mean = `r national_stats$post_m_prop_sat * 100`%, 95% credible interval (CI): `r national_stats$LCL_prop_sat * 100` - `r national_stats$UCL_prop_sat * 100`%) functioned as net N~2~O sinks during the summer months. The Western Mountains ecoregion had the smallest proportion of lakes predicted as N~2~O sinks (posterior mean = `r ecoreg_stats %>% filter(WSA9 == "WMT") %>% pull(post_m_prop_sat) * 100`%, 95% CI: `r ecoreg_stats %>% filter(WSA9 == "WMT") %>% pull(LCL_prop_sat) * 100` - `r ecoreg_stats %>% filter(WSA9 == "WMT") %>% pull(UCL_prop_sat) * 100`%) and the Northern Plains had the greatest (posterior mean = `r ecoreg_stats %>% filter(WSA9 == "NPL") %>% pull(post_m_prop_sat) * 100`%, 95% CI: `r ecoreg_stats %>% filter(WSA9 == "NPL") %>% pull(LCL_prop_sat) * 100` - `r ecoreg_stats %>% filter(WSA9 == "NPL") %>% pull(UCL_prop_sat) * 100`%), although the 95% credible intervals of the predicted distributions overlap across all ecoregions **(Table 1)**.

@@

The estimated extent of N~2~O undersaturation in CONUS lakes was surprisingly high, particularly for nitrogen rich ecoregions that are managed for agricultural production. This finding runs counter to the long-standing paradigm that surface waters are a source of N~2~O to the atmosphere [@Mosier_etal_1998] and joins other recent reports that lakes can function as N~2~O sinks. Webb et al. [@Webb_etal_2019] found that 67% of sampled reservoirs in an agricultural region of Canada were undersaturated with N~2~O during the summer months. An investigation of boreal systems in Canada found that 40% of rivers, lakes, and ponds sampled during the summer were functioning as N~2~O sinks[@Soued_etal_2016]. Similarly, 29% of surveyed boreal lakes in Finland were found to be N~2~O sinks during the summer [@Kortelainen_etal_2020]. While these latter surveys were conducted at regional scales, the national-scale results presented here provide the strongest evidence to date that most lakes function as N~2~O sinks during the summer months. This pattern held across all 9 major ecoregions in the CONUS, despite the tremendous variation in climate, geology, waterbody morphology, and land use.

@@

Despite widespread N~2~O undersaturation across the POI, estimated means for N~2~O saturation ratio (ratio of measured to equilibrium concentration) were indistinguishable from a value of one in all ecoregions **(Table 1, Fig. 2)**. The apparent contradiction of widespread N~2~O undersaturation and average dissolved N~2~O concentrations close to equilibrium can be attributed to the prominent right skew in the predicted distribution of dissolved N~2~O concentrations across the POI **(Fig. 2)**. Across the CONUS, and within each of the nine major ecoregions, the predicted distribution of N~2~O saturation ratios had a median value less than one, indicating that most lakes were undersaturated in these populations. However, the long right tails on these distributions, comprised of relatively rare lakes with very high levels of N~2~O supersaturation, were sufficient to drive the population means to values close to, or even greater than one. Thus, most lakes in the CONUS were functioning as N~2~O sinks during the summer months, but a few functioned as strong N~2~O sources, particularly at high NO~3~ and high water temperature **(Fig. 3)**.

@@

Lake size was also related to N~2~O saturation ratio, but the relationship was dependent on NO~3~ concentration **(Fig. 3)**.  Small lakes with low NO~3~ levels were predicted to be undersaturated, whereas small lakes with high NO~3~ were predicted to be highly supersaturated. Large lakes, regardless of NO~3~ status, were predicted to have N~2~O saturation ratios closer to equilibrium. . Thus N~2~O saturation ratio tended to deviate from a value of one more strongly in small lakes as compared to larger lakes.  This pattern can also be seen by calculating the absolute magnitude of the difference between the measured and equilibrium N~2~O concentration, known as $\Delta$ N~2~O.  When pooled across all ecoregions, $\Delta$ N~2~O is predicted to decline as lake size increases **(Fig. 4)**. This pattern holds regardless of whether N~2~O was supersaturated or undersaturated. Lake size and shape can affect dissolved gas concentration through several mechanisms. For example, N~2~O inputs from streams or groundwater[@Wang_etal_2023] may constitute a substantial N~2~O subsidy to small lakes but are more likely be diluted in larger lakes with greater volumes. Small lakes also tend to have lower water volume to benthic surface area ratios,and therefore stronger connections between sediment, an area of active N~2~O processing, and the air-water interface where our samples were collected. Furthermore, sediment may warm faster in small and shallow lakes, whereas thermal stratification may substantially delay sediment warming in large and deep waterbodies, resulting in greater rates of N~2~O production or consumption in small lakes. Thermal stratification may also prevent the exchange of water between the lake surface and the hypolimnion, thereby preventing N~2~O depleted or enriched water from advecting to the lake surface [@Beaulieu_etal_2015; @Beaulieu_etal_2014].

Persistent patterns in the air-water gas exchange rate across lake size classes may also contribute to greater N~2~O disequilibrium in smaller lakes. Air-water gas exchange will vent excess N~2~O from surface waters or facilitate the invasion of air that can replenish depleted N~2~O in undersaturated lakes. Regardless of whether a lake is supersaturated or undersaturated in N~2~O, air-water gas exchange will function to move the dissolved N~2~O concentration toward equilibrium. One of the primary drivers of air-water gas exchange in lentic waters is wind speed [@Vachon_etal_2013]. Wind speeds tend to be greater in large lakes with long fetches and lower in small lakes, particularly if they are sheltered by shoreline vegetation. Thus, greater air-water gas exchange rates in large lakes may maintain dissolved N~2~O concentrations close to equilibrium, whereas the effect of biogeochemistry on dissolved N~2~O may be more pronounced in small lakes because low air-water gas exchange limits the rate at which excess N~2~O can be ventilated or depleted N~2~O can be replenished. While previous surveys have documented relationships between lake size and dissolved methane concentration [@Pighini_etal_2018; @Juutinen_etal_2009; @Borges_etal_2011], to the best of our knowledge this is the first study to document a relationship between lake size and $\Delta$ N~2~O.

@@

![Figure 5: A) Predicted N~2~O emission rate and B) flux plotted against waterbody size for all 465,896 lakes in the population of interest. Points and bars reflect the posterior means (points) and 95% credible intervals, respectively, based on 2000 draws from the posterior predictive distribution for each waterbody. C) A comparison of the estimated relative flux contributions from lakes in each size class to the total CONUS flux. The red line connects the mean of the posterior estimate of the ratio, at each size class, based on 2000 draws. The black polygon indicates the 95% credible intervals for the ratio estimates across the size classes.](manuscript_figures/n2oFluxAndEmissionRateVsContinuousArea.tiff)

@@

Despite the estimated widespread N~2~O undersaturation among CONUS lakes, the credible intervals for ecoregion-level flux estimates included zero for all but three ecoregions **(Table 1)**. Thus, status as a net source versus sink of N~2~O was uncertain for most ecoregions. However, the estimates for the Southern Appalachians, Southern Plains, and Temperate Plains included only positive values, suggesting they were likely to have functioned as net N~2~O sources.  We were also unable to infer any clear differences in total flux between ecoregions. This result may sound surprising given the large differences in environmental conditions across these ecoregions, but it should not be mistaken as inferring "no differences" as it primarily emphasizes the challenges in making sufficiently precise estimates at this scale; even with unprecedented data coverage, a robust survey methodology, and relatively advanced modeling techniques.

The estimate for flux across the POI was centered over positive values (posterior mean = `r national_stats %>% select(post_m_fn2o) %>% pull() %>% round(., 0)` metric tons N~2~O year^-1^); however, the credible interval suggested a good deal of uncertainty and included zero (`r national_stats %>% select(LCL_fn2o) %>% pull() %>% round(., 0)` - `r national_stats %>% select(UCL_fn2o) %>% pull() %>% round(., 0)` metric tons N~2~O year^-1^). Again, we could not confidently infer the extent to which CONUS lakes, overall, were a net source or sink of N~2~O. Lauerwald et al [@Lauerwald_etal_2019] used a mechanistic model to estimate that CONUS lakes emit 8,425 metric tons N~2~O year^-1^.  That estimate is nearly 25-fold greater than the posterior mean of our estimate (344 metric tons N~2~O year^-1^), despite our estimate including an order of magnitude more lakes than Lauerwald et al.[@Lauerwald_etal_2019]  A possible explanation for the large difference between these estimates is that the Lauerwald model does not allow for lakes to behave as N~2~O sinks, a limitation that extends to many mechanistic models used to estimate N~2~O fluxes from inland waters[@Marzadri_etal_2021;@Marzadri_etal_2017;@Mosier_etal_1998]]. However, even models that allow for net N~2~O uptake can overpredict N~2~O flux. For example, Wang et al.[@Wang_etal_2023] used a model that allows for N~2~O consumption via denitrification to predict that CONUS lakes emitted an average of 47,614 metric tons N~2~O year^-1^ from 2011-2015.  This estimate is two orders of magnitude greater than the posterior mean of our estimate and suggests the mechanistic model does not properly describe N~2~O consumption.  Given our estimate of widespread N~2~O consumption in CONUS lakes, failure to properly describe this dynamic in aquatic N~2~O models may yield high-biased emission estimates.

@@

Although lakes in the largest lake size class tend to have N~2~O levels close to equilibrium, they emit more N~2~O than the other size classes due to their expansive surface area. The N~2~O flux from these lakes is also more uncertain than the smaller lake size classes because relatively small uncertainties in dissolved and equilibrium N~2~O concentrations translate to large uncertainties in flux when propagated across the surface of large lakes. Uncertainty in the national flux estimate was too large to determine if CONUS lakes functioned as a net source or sink of N~2~O. While our sample included some of the largest lakes in the CONUS, efforts to improve the accuracy of the lentic N~2~O flux estimate should include an assessment of intra-waterbody spatial variation in N~2~O, which can be important in large waterbodies[@Salk_etal_2019; @Borges_etal_2023], as well as careful accounting of measurement and analytical uncertainty, which can propagate to uncertainties in flux estimates.

@@

![SI Figure 1. Distribution of N~2~O emission rates by ecoregion. Each blue line (n=500) represents the distribution of values from a single realization of the posterior prediction.](manuscript_figures/SIfigure1.tiff)

>>>>

**Lakes, ponds, and reservoirs are estimated to be globally important sources of nitrous oxide (N~2~O) to the atmosphere but recent evidence of N~2~O uptake across a broad range of lakes have called the accuracy of emission estimates into question. Here we use a new national-scale dataset on dissolved N~2~O and a Bayesian hierarchical model to predict N~2~O concentration and emission rates in `r sframe %>% summarize(nlake = n()) %>% pull() %>% format(., big.mark=",", scientific=FALSE)` waterbodies in the conterminous U.S. (CONUS). We found that N~2~O undersaturation was widespread through the CONUS, with an estimated `r national_stats$post_m_prop_sat * 100`% (95% credible interval: `r national_stats$LCL_prop_sat * 100` - `r national_stats$UCL_prop_sat * 100`%) of lakes functioning as N~2~O sinks. The model predicts dissolved N~2~O concentrations reasonably well based partly on interactions between nitrate concentration, waterbody surface area, and water temperature. Despite working with the largest aquatic N~2~O dataset to date, our national-scale estimate of N~2~O emissions from CONUS lakes is poorly constrained, with a 95% credible interval ranging from net uptake to net emission (`r national_stats %>% select(LCL_fn2o) %>% pull() %>% round(., 0)` - `r national_stats %>% select(UCL_fn2o) %>% pull() %>% round(., 0)` metric tons N~2~O year^-1^). Widespread N~2~O undersaturation in CONUS waterbodies and a national-scale emission estimate that is too uncertain to determine if CONUS lakes are a net source or sink of N~2~O highlight the need to revisit N~2~O models which presume surface waters are a N~2~O source.**

@@

During the summer of 2017 the United States Environmental Protection Agency conducted the first national-scale survey of N~2~O concentrations and emissions in CONUS lakes as part of the National Lakes Assessment (NLA) [@USEPA_NLA17_2022].  The population of interest (POI) for the survey was `r sframe %>% summarize(nlake = n()) %>% pull() %>% format(., big.mark=",", scientific=FALSE)` CONUS lakes that had a surface area of at least 1 ha.  A representative sample (n=`r dg %>% filter(sample.source == "DG", visit.no == 1, sitetype == "PROB") %>% nrow()`) was selected from the POI using a generalized random-tessellation stratified survey design [@Stevens_Olsen_2004]**(Fig. 1)**. The survey design was stratified by state and used unequal probability categories based on lake size (surface area) to ensure that large lakes, which are far less common than small lakes, were included in the study.  Sample sites were distributed across all nine major CONUS ecoregions [@Herlihy_etal_2008] and varied widely in water chemistry, morphometry, watershed land-use, and climate [@epa_data]. The sample data and survey design information (e.g., stratification variables and unequal probability categories) were used to develop a Bayesian hierarchical model to predict dissolved and equilibrium N~2~O concentrations for the POI. Predictor variables for the model were restricted to those required to generate predictions that reproduced the observed N~2~O measures in an out-of-sample dataset and were available for the POI.  See Methods and SI for a detailed discussion of variable selection.

`r round(((dg %>% dplyr::filter(n2o.src.snk == "sink") %>% {nrow(.)} / dg %>% distinct(site.id) %>% {nrow(.)}) * 100), 0) %>% xfun::numbers_to_words(., cap = TRUE)` percent of the sampled lakes had dissolved N~2~O concentrations that were lower than the concentration expected if the lake were in equilibrium with the atmosphere, suggesting those lakes were functioning as N~2~O sinks **(Fig. 1)**. N~2~O concentrations were close to equilibrium values in some of these lakes, however, and after accounting for measurement uncertainty in both dissolved and equilibrium N~2~O concentration (See SI), we can confidently report that a minimum of `r round(((dg %>% dplyr::filter(n2o.src.snk.error == "sink") %>% {nrow(.)} / dg %>% distinct(site.id) %>% {nrow(.)}) * 100), 0)` percent of the sampled lakes were functioning as sinks, but the number could be as high as `r round(((dg %>% dplyr::filter(n2o.src.snk == "sink") %>% {nrow(.)} / dg %>% distinct(site.id) %>% {nrow(.)}) * 100), 0)` percent. When accounting for measurement error, `r round(((dg %>% dplyr::filter(n2o.src.snk.error == "undetermined") %>% {nrow(.)} / dg %>% distinct(site.id) %>% {nrow(.)}) * 100), 0)` percent had a dissolved N~2~O concentration that was indistinguishable from the equilibrium concentration and only `r round(((dg %>% dplyr::filter(n2o.src.snk.error == "source") %>% {nrow(.)} / dg %>% distinct(site.id) %>% {nrow(.)}) * 100), 0)` percent were supersaturated **(Fig. 1)**.

The Bayesian hierarchical model predicted that the majority lakes in the POI (posterior mean = `r national_stats$post_m_prop_sat * 100`%, 95% credible interval (CI): `r national_stats$LCL_prop_sat * 100` - `r national_stats$UCL_prop_sat * 100`%) functioned as net N~2~O sinks during the summer months. The Western Mountains ecoregion had the smallest proportion of lakes predicted as N~2~O sinks (posterior mean = `r ecoreg_stats %>% filter(WSA9 == "WMT") %>% pull(post_m_prop_sat) * 100`%, 95% CI: `r ecoreg_stats %>% filter(WSA9 == "WMT") %>% pull(LCL_prop_sat) * 100` - `r ecoreg_stats %>% filter(WSA9 == "WMT") %>% pull(UCL_prop_sat) * 100`%) and the Northern Plains had the greatest (posterior mean = `r ecoreg_stats %>% filter(WSA9 == "NPL") %>% pull(post_m_prop_sat) * 100`%, 95% CI: `r ecoreg_stats %>% filter(WSA9 == "NPL") %>% pull(LCL_prop_sat) * 100` - `r ecoreg_stats %>% filter(WSA9 == "NPL") %>% pull(UCL_prop_sat) * 100`%), although the 95% credible intervals of the predicted distributions overlap across all ecoregions **(Table 1)**.

@@

The estimated extent of N~2~O undersaturation in CONUS lakes was surprisingly high, particularly for nitrogen rich ecoregions that are managed for agricultural production. This finding runs counter to the long-standing paradigm that surface waters are a source of N~2~O to the atmosphere [@Mosier_etal_1998] and joins other recent reports that lakes can function as N~2~O sinks. Webb et al.[@Webb_etal_2019] found that 67% of sampled reservoirs in an agricultural region of Canada were undersaturated with N~2~O during the summer months and Kortelainen et al.[@Kortelainen_etal_2020] found that 29% of surveyed boreal lakes in Finland were N~2~O sinks during the summer. An investigation of boreal systems in Canada found that 40% of rivers, lakes, and ponds sampled during the summer were functioning as N~2~O sinks[@Soued_etal_2016] and all 34 lotic and lentic sites in the US National Ecological Observatory Network (NEON) functioned as N~2~O sinks at least once during 1.5 - 6 years of monitoring[@Aho_etal_2023]. While these latter surveys were conducted at regional scales or had relatively few observtion sites, the national-scale results presented here provide the strongest evidence to date that most lakes function as N~2~O sinks during the summer months. This pattern held across all 9 major ecoregions in the CONUS, despite the tremendous variation in climate, geology, waterbody morphology, and land use.

@@

Despite widespread N~2~O undersaturation across the POI, estimated means for N~2~O saturation ratio (ratio of measured to equilibrium concentration) were indistinguishable from a value of one in all ecoregions **(Table 1, Fig. 2)**. The apparent contradiction of widespread N~2~O undersaturation and average dissolved N~2~O concentrations close to equilibrium can be attributed to the prominent right skew in the predicted distribution of dissolved N~2~O concentrations across the POI **(Fig. 2)**. Across the CONUS, and within each of the nine major ecoregions, the predicted distribution of N~2~O saturation ratios had a median value less than one, indicating that most lakes in these populations were undersaturated . However, the long right tails on these distributions, comprised of relatively rare lakes with very high levels of N~2~O supersaturation, were sufficient to drive the population means to values close to, or even greater than one. Thus, most lakes in the CONUS were functioning as N~2~O sinks during the summer months, but a few functioned as strong N~2~O sources, particularly at high NO~3~ and high water temperature **(Fig. 3)**.

@@

Lake size was also related to N~2~O saturation ratio, but the relationship was dependent on NO~3~ concentration **(Fig. 3)**.  Small lakes with low NO~3~ levels were predicted to be undersaturated, whereas small lakes with high NO~3~ were predicted to be highly supersaturated. Large lakes, regardless of NO~3~ status, were predicted to have N~2~O saturation ratios closer to equilibrium. Thus N~2~O saturation ratio tended to deviate from a value of one more strongly in small lakes as compared to larger lakes.  This pattern can also be seen by calculating the absolute magnitude of the difference between the measured and equilibrium N~2~O concentration, known as $\Delta$ N~2~O.  When pooled across all ecoregions, $\Delta$ N~2~O is predicted to decline as lake size increases **(Fig. 4)**. This pattern holds regardless of whether N~2~O was supersaturated or undersaturated. Lake size and shape can affect dissolved gas concentration through several mechanisms. For example, N~2~O inputs from streams or groundwater[@Wang_etal_2023] may constitute a substantial N~2~O subsidy to small lakes but are more likely be diluted in larger lakes with greater volumes. Small lakes also tend to have lower water volume to benthic surface area ratios[@Cael_etal_2017], and therefore stronger connections between sediment, an area of active N~2~O processing, and the air-water interface where our samples were collected. Furthermore, sediment may warm faster in small and shallow lakes, whereas thermal stratification may substantially delay sediment warming in large and deep waterbodies, resulting in greater rates of N~2~O production or consumption in small lakes. Thermal stratification may also prevent the exchange of water between the lake surface and the hypolimnion, thereby preventing N~2~O depleted or enriched water from advecting to the lake surface [@Beaulieu_etal_2015; @Beaulieu_etal_2014].

Persistent patterns in the air-water gas exchange rate across lake size classes may also contribute to greater N~2~O disequilibrium in smaller lakes. Air-water gas exchange will vent excess N~2~O from surface waters or facilitate the invasion of air that can replenish depleted N~2~O in undersaturated lakes. Regardless of whether a lake is supersaturated or undersaturated in N~2~O, air-water gas exchange will function to move the dissolved N~2~O concentration toward equilibrium. One of the primary drivers of air-water gas exchange in lentic waters is wind speed [@Vachon_etal_2013]. Wind speeds tend to be greater in large lakes with long fetches and lower in small lakes, particularly if they are sheltered by shoreline vegetation. Thus, greater air-water gas exchange rates in large lakes may maintain dissolved N~2~O concentrations close to equilibrium, whereas the effect of biogeochemistry on dissolved N~2~O may be more pronounced in small lakes because low air-water gas exchange limits the rate at which excess N~2~O can be ventilated or depleted N~2~O can be replenished. While previous surveys have documented negative relationships between lake size and dissolved methane concentration [@Pighini_etal_2018; @Juutinen_etal_2009; @Borges_etal_2011], to the best of our knowledge this is the first study to document a relationship between lake size and $\Delta$ N~2~O.

@@

![Figure 5: A) Predicted N~2~O emission rate and B) flux plotted against waterbody size for all 465,896 lakes in the population of interest. Points (red) and bars (black) reflect the posterior means and 95% credible intervals, respectively, based on 2000 draws from the posterior predictive distribution for each waterbody. C) Proportion of total CONUS N~2~O flux attributed to each of the 465,896 lakes in the POI. Points (red) and bars (black) reflect the posterior means and 95% credible intervals, respectively, based on 2000 draws from the posterior predictive distribution for each waterbody](manuscript_figures/n2oFluxAndEmissionRateVsContinuousArea.tiff)

@@

Despite the estimated widespread N~2~O undersaturation among CONUS lakes, credible intervals for ecoregion-level flux estimates included zero for all but three ecoregions **(Table 1)**. Thus, status as a net source versus sink of N~2~O was uncertain for most ecoregions. However, the estimates for the Southern Appalachians, Southern Plains, and Temperate Plains included only positive values, suggesting they were likely to have functioned as net N~2~O sources.  We were also unable to infer any clear differences in total flux between ecoregions. This result may sound surprising given the large differences in environmental conditions across these ecoregions, but it should not be mistaken as inferring "no differences" as it primarily emphasizes the challenges in making sufficiently precise estimates at this scale; even with unprecedented data coverage, a robust survey methodology, and relatively advanced modeling techniques.

The estimate for flux across the POI was centered over positive values (posterior mean = `r national_stats %>% select(post_m_fn2o) %>% pull() %>% round(., 0)` metric tons N~2~O year^-1^); however, the credible interval suggested a good deal of uncertainty and included zero (`r national_stats %>% select(LCL_fn2o) %>% pull() %>% round(., 0)` - `r national_stats %>% select(UCL_fn2o) %>% pull() %>% round(., 0)` metric tons N~2~O year^-1^). Again, we could not confidently infer the extent to which CONUS lakes, overall, were a net source or sink of N~2~O. Lauerwald et al [@Lauerwald_etal_2019] used a mechanistic model to estimate that CONUS lakes emit 8,425 metric tons N~2~O year^-1^.  That estimate is nearly 25-fold greater than the posterior mean of our estimate (344 metric tons N~2~O year^-1^), despite our estimate including an order of magnitude more lakes than Lauerwald et al.[@Lauerwald_etal_2019]  A possible explanation for the large difference between these estimates is that the Lauerwald model does not allow for lakes to behave as N~2~O sinks, a limitation that extends to many mechanistic models used to estimate N~2~O fluxes from inland waters[@Marzadri_etal_2021;@Marzadri_etal_2017;@Mosier_etal_1998]. However, even models that allow for net N~2~O uptake can overpredict N~2~O flux. For example, Wang et al.[@Wang_etal_2023] used a model that allows for N~2~O consumption via denitrification to predict that CONUS lakes emitted an average of 47,614 metric tons N~2~O year^-1^ from 2011-2015.  This estimate is two orders of magnitude greater than the posterior mean of our estimate and suggests the mechanistic model does not properly describe N~2~O consumption.  Given our estimate of widespread N~2~O consumption in CONUS lakes, failure to properly describe this dynamic in aquatic N~2~O models may yield high-biased emission estimates.

@@

Although lakes in the largest lake size class tend to have N~2~O levels close to equilibrium, they make a large contribution to CONUS N~2~O flux due to their expansive surface area. The N~2~O flux from large lakes is poorly constrained, however, because relatively small uncertainties in dissolved and equilibrium N~2~O concentrations translate to large uncertainties in flux when propagated across the surface of large lakes. While our sample included some of the largest lakes in the CONUS, efforts to improve the accuracy of the lentic N~2~O flux estimate should include an assessment of intra-waterbody spatial variation in N~2~O, which can be important in large waterbodies[@Salk_etal_2019; @Borges_etal_2023], as well as careful accounting of measurement and analytical uncertainty, which can propagate to uncertainties in flux estimates. The measurement methods employed in this study have a 95% credible interval of 0.11 (unitless) on the N~2~O saturation ratio. The 95% credible interval can be halved by employing gas chromatography optimized for quantifying N~2~O in the range of values observed in this study [@Hall_2024] and decreased by an order of magnitude using membrane inlet mass spectrometry[@Genthner_etal_2013; @Speir_etal_2023].

@@

### 2025-02-18 scripts/dataMunge.Rmd
Completed revisions in response to Lester and Jana's tech review.

@@

```{r}
# Read dissolved gas data file
dg <- read.table(file = paste0(localPath, 
                               "/Environmental Protection Agency (EPA)/",
                               "ORD NLA17 Dissolved Gas - Documents/",
                               "inputData/nla17gasDataAggregated_2021-02-01.txt"),
                 header = TRUE, sep = "\t", as.is = TRUE) %>%
  filter(!grepl("AK", site.id)) # omit Alaska sites
```

## Manipulate
### Implement unit conversion and create source/sink column:
``` {r}
dg <- dg %>%
  # unit conversion
  mutate(dissolved.ch4.umol = dissolved.ch4 * 1000000, # mol/L -> umol/L
         dissolved.co2.umol = dissolved.co2 * 1000000, # mol/L -> umol/L,
         dissolved.n2o.nmol = dissolved.n2o * 1000000000, # mol/L -> nmol/L
         sat.ch4.umol = sat.ch4 * 1000000, # mol/L -> umol/L
         sat.co2.umol = sat.co2 * 1000000, # mol/L -> umol/L
         sat.n2o.nmol =  sat.n2o * 1000000000, # mol/L -> nmol/L
         # add source/sink column
         co2.src.snk = ifelse(co2.sat.ratio > 1, "source", "sink"),
         ch4.src.snk = ifelse(ch4.sat.ratio > 1, "source", "sink"),
         n2o.src.snk = ifelse(n2o.sat.ratio > 1, "source", "sink")) %>%
  # remove fields no longer needed
  select(-dissolved.ch4, -dissolved.co2, -dissolved.n2o,
         -sat.ch4, -sat.co2, - sat.n2o)
```

>>>>

Roy Martin modeled the uncertainty in the N2O saturation ratio in DG_sensitivity_to_measurement_error.Rmd (Saturation ratio/Standard GC sections). The absolute error in the simulated values were written to desk and read in below. These error estimates are used to define uncertainty in the N2O saturation ratio.

@@

```{r}
# Read dissolved gas data file
dg <- read.table(file = paste0(localPath, 
                               "/Environmental Protection Agency (EPA)/",
                               "ORD NLA17 Dissolved Gas - Documents/",
                               "inputData/nla17gasDataAggregated_2021-02-01.txt"),
                 header = TRUE, sep = "\t", as.is = TRUE) %>%
  filter(!grepl("AK", site.id)) # omit Alaska sites

# Read simulated saturation ratio values (See DG_sensitivity_to_measurement_error.Rmd, Saturation ratio/Standard GC)
S_a_me_abs <- readRDS(paste0(localPath, 
                             "/Environmental Protection Agency (EPA)/",
                             "ORD NLA17 Dissolved Gas - Documents/",
                             "inputData/S_a_me_abs.RDS"))

## Manipulate
### Implement unit conversion and create source/sink column:
``` {r}
dg <- dg %>%
  # unit conversion
  mutate(dissolved.ch4.umol = dissolved.ch4 * 1000000, # mol/L -> umol/L
         dissolved.co2.umol = dissolved.co2 * 1000000, # mol/L -> umol/L,
         dissolved.n2o.nmol = dissolved.n2o * 1000000000, # mol/L -> nmol/L
         sat.ch4.umol = sat.ch4 * 1000000, # mol/L -> umol/L
         sat.co2.umol = sat.co2 * 1000000, # mol/L -> umol/L
         sat.n2o.nmol =  sat.n2o * 1000000000, # mol/L -> nmol/L
         # add source/sink column
         co2.src.snk = ifelse(co2.sat.ratio > 1, "source", "sink"),
         ch4.src.snk = ifelse(ch4.sat.ratio > 1, "source", "sink"),
         n2o.src.snk = ifelse(n2o.sat.ratio > 1, "source", "sink"),
         # recompute using error from Roy's uncertainty simulations
         n2o.src.snk.error = case_when(
           abs(n2o.sat.ratio - 1) < quantile(S_a_me_abs, probs = c(0.975)) ~ "undetermined",
           TRUE ~ n2o.src.snk)) %>%
  # remove fields no longer needed
  select(-dissolved.ch4, -dissolved.co2, -dissolved.n2o,
         -sat.ch4, -sat.co2, - sat.n2o)
```
